Fig3

### 2020-04-13 analysis/figure-3.Rmd
tables with prl counts
new DEG table

candidategenesslim <- c("OXT", "AVP", "GNRH1", "GNRHR", 
                    "AR",  "CYP19A1", 
                     "AVPR1A", "AVPR1B", "AVPR2","VIP",
                  "DRD1", "DRD2", 
                  "PRL", "PRLR",  
                    "ESR1","ESR2", "LBH",

                    "FOS", "JUN", "EGR1", "BDNF"
                      )

# all candidate genes: AR, AVP, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, EGR1, ESR1, ESR2, FOS, GNRH1, GNRHR, JUN, LBH, OXT, PRL, PRLR, VIP

# DEG candidate genes: AR, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, ESR1, FOS, GNRHR, LBH, PRL, PRLR,

# Candidate genes that were not differentially expressed: AVP, AVPR1B, ESR2,  GNRH1, JUN, OXT, VIP

@@

```{r }
geneids <- read_csv("../metadata/00_geneinfo.csv")

@@

hypvsd <- getcandidatevsd(candidategenesslim, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(candidategenesslim, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(candidategenesslim, "gonad", sexlevels)
head(hypvsd)
head(pitvsd)
head(gonvsd)

@@

```{r fig3, fig.width=7, fig.height=7}
makeboxplots <- function(df, whichgene, mysubtitle, whichsex, whichtimepoint){
  p <- df %>%
    filter(treatment %in% charlevels,
           gene %in% whichgene,
           sex %in% whichsex) %>%
    filter(treatment %in% whichtimepoint) %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot(outlier.shape = NA) + 
    geom_jitter(size = 0.5, width = 0.1) +
    facet_wrap(~gene, scales = "free_y", nrow = 1) +
    theme_B3() +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme(#axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(y = "gene expression" , x = "Sequential stages with candidate DEGs", subtitle = mysubtitle) +
    theme(strip.text = element_text(face = "italic"))
  return(p)
}

## hyp 
a <- makeboxplots(hypvsd, c("BDNF","CYP19A1", "DRD1", "EGR1"), 
                  "Female hypothalamus", "female", c("hatch", "n5"))  +
  theme(axis.title.x = element_blank())

b <- makeboxplots(hypvsd, c("AR"),"Male hypothalamus", "male", 
                  c( "inc.d9", "inc.d17"))  +
  theme(axis.title = element_blank())

ab <- plot_grid(a,b, rel_widths = c(4,1),  labels = "auto", label_size = 8)

c <- makeboxplots(pitvsd, c("ESR1","GNRHR"), "Female pituitary", "female", c("bldg", "lay", "inc.d3"))   +
  theme(axis.title.x = element_blank())
d <- makeboxplots(pitvsd, c("LBH", "PRL" ), " ", "female", c("inc.d9", "inc.d17", "hatch", "n5"))  +
  theme(axis.title  = element_blank())
e <- makeboxplots(pitvsd, c( "AVPR2"), " ", "female", c( "hatch", "n5"))  +
  theme(axis.title  = element_blank())

cde <- plot_grid(c,e,d, rel_widths = c(6,2.25,7), nrow = 1,  labels = c("c"), label_size = 8)

g <- makeboxplots(gonvsd, c( "AVPR1A"), "Female gonads", "female", c( "lay", "inc.d3", "inc.d9"))  +
  labs(caption = " ", x = " ")
h <- makeboxplots(gonvsd, c( "EGR1", "FOS", "PRLR"), " ", "female", c( "lay", "inc.d3"))  +
  theme(axis.title.y = element_blank()) + labs(caption = " ")
f <- makeboxplots(pitvsd, c("LBH", "PRL"), "Male pituitary", "male", 
                  c("inc.d9", "inc.d17")) + 
  theme(plot.caption = element_text(face = "italic")) + 
  labs(x = " ",
       caption = "Candidate genes that were not differentially expressed from bldg to n5: AVP, AVPR1B, ESR2,  GNRH1, JUN, OXT, VIP.")

ghf <- plot_grid(g,h,f, nrow = 1, rel_widths = c(3,6,4),  labels = c("d", " ", "e"), label_size = 8)

plot_grid(ab, cde, ghf, nrow = 3, rel_heights = c(1,1,1.25))

@@

```{r writefiles}
write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
```

>>>>

candidategenesslim <- c("OXT", "AVP", "GNRH1", "GNRHR", 
                    "AR",  "CYP19A1", 
                    "AVPR1A", "AVPR1B", "AVPR2","VIP",
                    "DRD1", "DRD2", 
                    "PRL", "PRLR",  
                    "ESR1","ESR2", "LBH",  
                    "FOS", "JUN", "EGR1", "BDNF")

# candidategenesslim: AR, AVP, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, EGR1, ESR1, ESR2, FOS, GNRH1, GNRHR, JUN, LBH, OXT, PRL, PRLR, VIP
# DEG candidategenesslim: AR, AVPR1A, AVPR1B, AVPR2, BDNF, CYP19A1, DRD1, ESR1, FOS, GNRHR, LBH, PRL, PRLR, 
# non-DEG candidategenesslim: AVP, AVPR1B, ESR2,  GNRH1, JUN, OXT, VIP

hormonepapergenes <- c("GNRH1", "CGNRH-R", "GNRHR",  
                       "FSHB", "FSHR", "LHCGR", # no LSH
                       "VIP", "VIPR1",
                       "ESR1", "AR", "CYP19A1", "HSD17B1", # receptors synthesis
                       "PGR" , "HSD3B2", 
                       "PRL", "PRLR")
```

@@

```{r }
geneids <- read_csv("../metadata/00_geneinfo.csv") %>% select(-X1)

@@

hypvsd <- getcandidatevsd(hormonepapergenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(hormonepapergenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(hormonepapergenes, "gonad", sexlevels)
head(hypvsd)
tail(gonvsd)

@@

```{r DEGs}
# summary DEG results from DESeq2
candidateDEGS <- read_csv("../../musicalgenes/data/allDEG.csv") %>%
  select(-X1) %>%
  filter(gene %in% hormonepapergenes)
head(candidateDEGS)

# get genes to plot
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "gonad", sex == "female", comparison != "control_bldg")

# summarize all differences
candidateDEGtable <- candidateDEGS %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-")) %>%
  mutate(geneposneg = paste(gene, posneg, sep = "")) %>%
  select(sex, tissue, comparison, geneposneg) %>%
  group_by(sex, tissue, comparison) %>%
   arrange(geneposneg) %>%
  summarize(genes = str_c(geneposneg, collapse = " ")) %>% 
  pivot_wider(names_from = comparison, values_from = genes)  %>% 
  ungroup(tissue) %>%
  mutate(tissue = factor(tissue, levels =  tissuelevel))  %>% 
  arrange(sex, tissue)  %>% 
  select(sex, tissue, bldg_lay, lay_inc.d3, inc.d9_inc.d17, hatch_n5, control_bldg)
candidateDEGtable

newboxplot <- function(df, mygenes, whichsex, whichstage){

  p <- df %>%
    filter(gene %in% mygenes, 
           sex %in% whichsex,
           treatment %in% whichstage) %>%
    droplevels() %>%
    mutate(tissue = factor(tissue, levels = tissuelevel)) %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot() +
    geom_jitter(size = 0.5, width = 0.1) +
    facet_grid(tissue~gene, scales = "free_y") +
    theme_B3() +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text.x = element_text(face = "italic"),
          axis.title.x = element_blank())  +
    labs(x = "Parental stage",
         y = "variance stabilized expression")
  return(p)
}

a1 <- newboxplot(hypvsd, c("PGR",  "CYP19A1", "LHCGR" ), "female", c( "hatch", "n5")) + 
  theme(strip.text.y = element_blank()) + labs(subtitle = "Females")
a2 <- newboxplot(hypvsd, c("AR"), "male", c( "hatch", "n5")) + 
  theme(axis.title.y = element_blank()) + labs(subtitle = "Males")

b1 <- newboxplot(gonvsd, c("HSD3B2"), "female", c( "bldg", "lay"))  +
  theme(strip.text.y = element_blank(),
        axis.title.y = element_blank()) + labs(subtitle = "Females")
b2 <- newboxplot(gonvsd, c("VIPR1",  "PRLR" ), "female", c( "lay", "inc.d3"))  + 
  theme(axis.title.y = element_blank()) + labs(subtitle = " ")

ab <- plot_grid(a1,a2,b1,b2, rel_widths = c(6,2.5,2,4), labels = c("a", " ", "b", " "), label_size = 8, nrow = 1)

c1 <- newboxplot(pitvsd, c( "ESR1", "GNRHR"), "female", c("bldg", "lay",  "inc.d3")) 
c2 <- newboxplot(pitvsd, c( "PRL"), "female", c("inc.d9", "inc.d17", "hatch", "n5")) 
c3 <- newboxplot(pitvsd, c("VIPR1",  "PRL" ), "male", c( "inc.d9", "inc.d17"))

c <- plot_grid(c1 + theme(strip.text.y = element_blank()) + labs(subtitle = "Females"),
          c2 + theme(strip.text.y = element_blank(),
                    axis.title.y = element_blank()) + labs(subtitle = " ") ,
          c3 + theme(axis.title.y = element_blank()) + labs(subtitle = "Males"), 
          nrow = 1, rel_widths = c(6,4,4),
          labels = c("c"), label_size = 8)

plot_grid(ab,c, nrow = 2)

@@

```{r writefiles}
#write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
#write.csv(candidatevsd, "../results/candidatecounts.csv")
write.csv(candidateDEGtable, "../results/candidateDEGtable.csv")
